---
title: "GV300 Showcase: Analysing the 2024 UK General Election"
description: |
  A showcase of my best code and visualisation from GV300, exploring
  constituency sizes, voter turnout, and representation in the 2024 UK
  General Election.
author:
  - name: Gabdolgalim Satpayev
date: 2026-02-15
output:
  distill::distill_article:
    toc: true
    code_folding: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

The 2024 UK General Election provides a rich dataset for examining how constituency sizes vary across the four nations, and how voter turnout differs between regions. In this showcase I present my best data-wrangling and visualisation code from GV300 Homework 2, where I analysed official candidate-level results published by the UK Parliament.

The analysis addresses three questions:

1. How much do constituency sizes vary, and what does that mean for equal representation?
2. Which visualisation best reveals cross-country differences?
3. Who isn't voting, and where is turnout lowest?

## Data Preparation

The code below loads the official 2024 candidacy-level CSV, cleans variable names, aggregates to constituency level, and computes a turnout measure.

```{r data-prep}
library(dplyr)
library(ggplot2)
library(stringr)

uk_csv <- "https://electionresults.parliament.uk/general-elections/6/candidacies.csv"
uk_raw <- read.csv(uk_csv, stringsAsFactors = FALSE)

# Standardise column names to lower_snake_case
names(uk_raw) <- names(uk_raw) %>%
  tolower() %>%
  str_replace_all("\\.", "_")

# Remove by-elections and keep key variables
uk1 <- uk_raw %>%
  filter(election_is_by_election == "false")

uk2 <- uk1 %>%
  select(
    constituency = constituency_name,
    country      = country_name,
    valid_votes  = candidate_vote_count,
    electorate   = electorate
  )

# Aggregate from candidate-level to constituency-level
uk3 <- uk2 %>%
  group_by(country, constituency) %>%
  summarise(
    valid_votes = sum(valid_votes, na.rm = TRUE),
    electorate  = max(electorate, na.rm = TRUE),
    .groups     = "drop"
  )

# Compute turnout and drop invalid rows
uk <- uk3 %>%
  mutate(turnout_registered = 100 * valid_votes / electorate) %>%
  filter(is.finite(turnout_registered), electorate > 0)

knitr::kable(
  head(uk, 10),
  caption = "First 10 constituencies after cleaning",
  digits = 1
)
```

After cleaning, the dataset contains **`r nrow(uk)` constituencies** across `r length(unique(uk$country))` countries of the United Kingdom.

## Smallest and Largest Constituencies

```{r extremes}
smallest <- uk %>%
  arrange(valid_votes) %>%
  slice_head(n = 5) %>%
  select(constituency, country, valid_votes, electorate, turnout_registered)

largest <- uk %>%
  arrange(desc(valid_votes)) %>%
  slice_head(n = 5) %>%
  select(constituency, country, valid_votes, electorate, turnout_registered)

knitr::kable(smallest, caption = "Five smallest constituencies by valid votes", digits = 1)
knitr::kable(largest, caption = "Five largest constituencies by valid votes", digits = 1)

ratio <- max(uk$valid_votes) / min(uk$valid_votes)
```

The largest constituency has roughly **`r round(ratio, 1)` times** as many valid votes as the smallest. Because each MP carries one vote in Parliament regardless of constituency size, voters in smaller constituencies effectively have more representation per person — raising questions about equal representation.

## Visualisation: Cross-Country Comparison

A faceted histogram is a natural first attempt, but it is hard to compare across panels with different y-scales. A **boxplot** makes cross-country differences immediately visible on a single axis.

```{r boxplot, fig.width=8, fig.height=5, layout="l-body-outset"}
p_box <- ggplot(uk, aes(x = country, y = valid_votes / 1000, fill = country)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  labs(
    title    = "Constituency Sizes by Country (2024 UK General Election)",
    subtitle = "Boxplots allow direct comparison of medians, spread, and outliers",
    x        = "Country",
    y        = "Valid Votes (thousands)",
    fill     = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

print(p_box)
```

## Summary Statistics by Country

```{r summary-stats}
size_summary <- uk %>%
  group_by(country) %>%
  summarise(
    n_constituencies = n(),
    mean_votes       = mean(valid_votes),
    median_votes     = median(valid_votes),
    sd_votes         = sd(valid_votes),
    min_votes        = min(valid_votes),
    max_votes        = max(valid_votes),
    .groups          = "drop"
  ) %>%
  arrange(desc(mean_votes))

knitr::kable(
  size_summary,
  col.names = c("Country", "N", "Mean", "Median", "SD", "Min", "Max"),
  caption   = "Constituency sizes (valid votes) by country",
  digits    = 0,
  format.args = list(big.mark = ",")
)
```

## Who Isn't Voting?

```{r non-voters}
non_voter_analysis <- uk %>%
  mutate(non_voters = electorate - valid_votes)

total_non_voters     <- sum(non_voter_analysis$non_voters)
total_electorate     <- sum(non_voter_analysis$electorate)
non_voter_percentage <- 100 * total_non_voters / total_electorate

top_non_voters <- non_voter_analysis %>%
  arrange(desc(non_voters)) %>%
  slice_head(n = 10) %>%
  select(constituency, country, electorate, valid_votes, non_voters, turnout_registered)

knitr::kable(
  top_non_voters,
  caption = "Top 10 constituencies by number of non-voters",
  digits  = 1,
  format.args = list(big.mark = ",")
)
```

Across all constituencies, approximately **`r format(round(total_non_voters), big.mark = ",")`** registered voters did not vote — about **`r round(non_voter_percentage, 1)`%** of the electorate. If non-voters are systematically different from voters (younger, poorer, more disengaged), then election outcomes may not reflect the preferences of the broader population.

## Turnout by Category and Country

```{r turnout-cats, fig.width=8, fig.height=5, layout="l-body-outset"}
turnout_categories <- uk %>%
  mutate(
    turnout_category = case_when(
      turnout_registered < 55 ~ "Low",
      turnout_registered < 65 ~ "Medium",
      turnout_registered < 75 ~ "High",
      TRUE                    ~ "Very High"
    ),
    turnout_category = factor(
      turnout_category,
      levels = c("Low", "Medium", "High", "Very High")
    )
  )

category_counts <- turnout_categories %>%
  group_by(turnout_category, country) %>%
  summarise(n_constituencies = n(), .groups = "drop")

p_cats <- ggplot(category_counts,
                 aes(x = turnout_category, y = n_constituencies, fill = country)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Constituencies by Turnout Category",
    x     = "Turnout Category",
    y     = "Number of Constituencies",
    fill  = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    legend.position = "bottom"
  )

print(p_cats)
```

## High vs Low Turnout Constituencies

```{r high-low}
high_turnout <- uk %>% filter(turnout_registered > 70)
low_turnout  <- uk %>% filter(turnout_registered < 60)

high_summary <- high_turnout %>%
  group_by(country) %>%
  summarise(
    n = n(),
    mean_electorate = mean(electorate),
    mean_turnout    = mean(turnout_registered),
    .groups = "drop"
  ) %>%
  mutate(group = "High (>70%)")

low_summary <- low_turnout %>%
  group_by(country) %>%
  summarise(
    n = n(),
    mean_electorate = mean(electorate),
    mean_turnout    = mean(turnout_registered),
    .groups = "drop"
  ) %>%
  mutate(group = "Low (<60%)")

comparison <- bind_rows(high_summary, low_summary) %>%
  select(group, country, n, mean_electorate, mean_turnout)

knitr::kable(
  comparison,
  col.names = c("Group", "Country", "N", "Mean Electorate", "Mean Turnout (%)"),
  caption   = "Comparison of high- and low-turnout constituencies",
  digits    = 1,
  format.args = list(big.mark = ",")
)
```

Differences in turnout likely reflect socio-economic factors (deprivation, education), demographics (age structure), and political factors such as competitiveness and campaign intensity. Areas with greater political disengagement tend to exhibit lower turnout than places where voters are more engaged and parties campaign harder.

## Conclusion

This analysis demonstrates the use of `dplyr` for data wrangling and `ggplot2` for multi-dimensional visualisation. The boxplot and grouped bar chart effectively communicate cross-country differences in constituency size and voter turnout across the 2024 UK General Election.

Source code for this site is available at [github.com/stasi1950/.github.io](https://github.com/stasi1950/.github.io).
